Combining neural networks and geostatistics for landslide hazard assessment of San Salvador Metropolitan Area, El Salvador \newline \newline Combinando redes neuronales y geoestad\'{i}stica para evaluaci\'{o}n de deslizamientos de tierra del \'{a}rea Metropolitana de San Salvador, El Salvador

### 2015-06-09 paper.tex
09 June 2015 12:56 Finally Version

Throughout the year the country experiences a tropical climate with two seasons, a dry season (November to April) and wet seasons (May - October). The climate of El Salvador is generally warm. In the dry season there is very little rainfall but during rainy seasons heavy showers take place. Periodically El Salvador is affected by tropical storms and occasionally by hurricanes.

@@

\section{Brief State of Art}
\label{sec:brief}
Since the pioneering work shown in \cite{Carrara1983403}, several mathematical and statistical models have been proposed to model landslide susceptibility: deterministic models (\cite{hessd-10-12643-2013},  \cite{doi:10.1080/19475705.2010.498151} and \cite{Neu2012511}) and probabilistic models (\cite{Bern198839}, \cite{Chung2003451} and \cite{doi:10.1080/01431160310001618734} ).

Popular classification models have been used such as logistic regression (\cite{akgun2012} and \cite{gaskill} and \cite{garcia2008}) , neural networks (\cite{garcia2010}, \cite{Melchiorre2011410},\cite{Zeng2001374}, \cite{Ermini2005327} and \cite{Yesilnacar2005251})  and support vector machines (\cite{ballabio2012support} and \cite{tien2012landslide}).

According to \cite{van2006landslide}, the magnitude of a possible
slide is difficult to foresee as it depends on the magnitude of the triggering event and the environmental conditions (e.g., height of water table) at the moment of the event. Because of these complex relationships between the dependent variable and causal factors, and since neural networks are particularly useful for complex non-linear relationships in large datasets, despite the disadvantages such as greater computational burden.

\section{Study area}
\label{sec:studyarea}
MASS is formed by 14 municipalities among them is located San Salvador, the capital of the country. It has 587 $km^{2}$ and according to the most recent census \cite{minecon} it has 1,566,629 inhabitans with a density of population of 2669 hab/km2. MASS is extended over a flat erosional surface (650-760 m.a.s.l.), gently sloping to the east towards Ilopango Caldera also known as Ilopango Lake. MASS is bordered to the south by the Balsamo Range and San Jacinto Hill, to the west by San Salvador volcano and Mariona Hills (refer to Fig. \ref{fig:mass01}). Human settlement is not restricted to the plain, spreads up to surrounding heights and volcano flanks.

All rocks exposed in MASS have volcanic origin and consists of intercaled primary and reworked deposits from Late Tertiary to Holocene age \cite{schmidt1975}. The younger ones are known as Tierra Blanca pyroclastic ash deposits and forms a relative continuous layer in majority of MASS with an average thickness of 4 m \cite{schmidt1975}. These are deposits derived mainly from Ilopango caldera eruptions where they present more than 50 m of thickness \cite{schmidt1975}. Tierra Blanca soils is highly susceptible to erosion, one of the most common dangerous effects of the erosion in this area are landslides during heavy rainfall and earthquake ground shaking \cite{schmidt1975}. Ch\'{a}vez et al. \cite{chavez2014a} identified high and moderate erosion hazard in the majority of MASS area. The more recent layer Tierra Blanca, known as Tierra Blanca Joven (TBJ) \cite{hernan2004} presents very poor geotechnical properties, especially when there is an increment of soil moisture (i.e. during rainy seasons), results in high instability susceptibility (\cite{chavez2014b} and \cite{rolo2004}). Several incised rivers and ravines cross MASS area, due the large erosion rates of the area, natural slopes of these watercourses are often close to vertical and can reach highs of more than 10 m. Taking into account this property of volcanic soils, usually vertical slopes are cut to urban and road construction. As Bommer \& Rodriguez \cite{bommer} identified in Central America, although such slopes may remain stable for years, they may become unstable, abruptly and totally under the action of heavy rainfall or seismic shaking. Hern\'{a}ndez \cite{hernan2004} carried out a detailed description of this process in TBJ, where these instabilities are very common. 
In study area fault trends are characterized by steep dip angles (65\textdegree - 90\textdegree). The main fault system is characterized by an east-west trend; this trend is responsible the steep northern slope of the El Balsamo Range.  The secondary systems are with north, northwest and northeast trends offset the main fault system and are more active in the present tectonic setting \cite{schmidt1975}. However, all the mentioned fault families appear active. Thus the several fault systems have developed at different times, but been repeatedly activated \cite{rymer1987}. Finally there are also several ringlike faults that can be related to subsurface collapse due to volcanic tectonic subsidence \cite{schmidt1975}.
Last eruption of San Salvador volcano was in 1917.  The majority of the magma's total volume (97\%) extruded from several vents in the Northwest flank of the volcano, outside of MASS. However a fraction of MASS is located under the proximal volcanic hazard zone according to Sofield \cite{sofield2004}. In the steep slopes  San Salvador volcano, especially of Picacho, the highest peak of the volcano, located in the eastern side presents a high lahar hazard that threats MASS \cite{major2004}. On the other hand one of the most destructive eruptions in the world history was 430 A.D. TBJ eruption produced by the Ilopango Caldera. According to Dull \cite{dull2004}, that eruption produced total devastation in whole MASS area.

\section{Landslide inventory and data sources of input variables}
\label{sec:landsinvet}
The Landslide inventory was developed by Ministry of Environment and Natural Resources (MARN Spanish acronym) and it was conducted in two stages. The first was carried out, based on fieldwork after the 2001 earthquakes in El Salvador. The second stage consisted of a visual inventory of places where identified through satellite image from IRS-C sensor (Indian) in the same year.

@@

\begin{enumerate}
\item {\bf Geomorphology:} it means landforms that result from lithospheric dynamics of geographic area. This variable was obtained in the project of national land use plan in El Salvador. It is unpublished geomorphologic cartography in digital format provided by MARN.

@@

\item {\bf Geology:} Description of the geology of the area, geological map of  El Salvador (scale 1:100,000) \cite{weber}.

@@

\item {\bf Road distance:} Distance in kilometers to the nearest road. Road map was obtained from national cartography created by Centro Nacional de Registros de la Rep\'{u}blica de El Salvador (CNR).

\item {\bf Fault distance:} Distance in kilometers to the nearest fault. This information was obtained from geological map of El Salvador (scale 1:100,000) \cite{weber}.

@@

\section{Artificial Neural Network model for discrete choice}
Logistic regression represents a neural network with 
a neuron in the hidden layer whose output is dichotomous. The following adapted form of the Multilayer Perceptron (MLP) may be used for modeling binary classification problems. In the equation (\ref{eq:ann1}) $x_{k,i}$ are the observed values in the ith input variables belonging to the kth case. The expressions $w_{j,k}$ and $\lambda_{j}$ showed in the equations (\ref{eq:ann1}) and (\ref{eq:ann3}) are the parameters of the model.

@@

In the equation (\ref{eq:ann3}) $p_{i}$ is the the predicting probability for a network with $ k^{*} $ input characteristics and $ j^{*} $ neurons. In the context of the present research, $p_{i} $ represents the probability of landslide ocurrence.

@@

In the equation (\ref{eq:scaling}) $z$ is the new variable transformed, $ \bar{x} $ and $\sigma_{x}$ the mean and the standard deviation of the input variable to transform.

The method used for estimating the parameters of the model was an Hybrid Method \cite{McNelis2005}. Using a package developed in the R statistical software \cite{rproject} specifically developed to obtain a good estimation of the parameters of the neural network model . The R package can be accessed from the following Web address:

@@

@@

In standard statistical problems, correlation can be estimated from a scatterplot, when several data pairs ${x, y}$ are available. The spatial correlation between two observations of a variable $z(s)$ at locations $s_{1}$ and $s_{2}$ cannot be estimated, only a single pair is available. To estimate spatial correlation
from observational data, it is necessary to make stationarity assumptions
before we can make any progress. One commonly used form of stationarity
is intrinsic stationarity, which assumes that the process that generated the
samples is a random function $Z(s)$ composed of a mean and residual \cite{bivand2008applied}:

@@

\begin{equation}
 E\left[Z(s_{0}) - \hat{Z(s_{0})}\right]^{2}
\label{eq:meankrige}
\end{equation}

\section{Results}
The data were randomly divided into three sub-samples. The first was used for the
estimation of the parameters of the neural network model (training set), the second was used to choose the model with more generalization capability (validation set), and the last was
used to evaluate how well the model is able to generalize data not used in the estimation process (test set).

To determine the number of neurons in the hidden layer, the method of trial and error was used, starting with few neurons and increasing progressively the number of neurons in the hidden layer. Table \ref{annresults}
summarizes how well the models fit on the training, validation and test sets.

@@

The model with 17 neurons in the hidden layer was chosen because it presents the best performance in the validation set. After that a logistic regression model was fitted as a basis of comparison. In the Table \ref{logisticresults} is presented logistic regression model results. Clearly, neural network model chosen has better performance than logistic regression model.

@@

In order to achieve the above steps, an R script was developed. This script can be downloaded in:

@@

First of all one thousand fifty georeferenced points with their probability of occurrence calculated using neural network, were selected randomly for the map generation process leaving the others for validation of the kriging model. The Exploratory analysis showed the presence of spatial correlation, this was done making scatter plot of pairs $Z(s_{i})$ and  $Z(s_{j})$, grouped according to their separation distance, also the data did not show the presence of anisotropic effect.

Regarding variogram estimation the spherical, gaussian and exponential models were proposed. The validation of these models on data were not taken into account in the process of estimating the variogram showed that the best model was the exponential, since it had an $ R^{2} = 0.42 $ against $ R^{2} = 0.40 $ and $ R^{2} = 0.39 $ of the models spherical and gaussian respectively. Using exponential model, spatial predictions were calculated on a grid of 1500 $\times$ 1500 and exported to Geotiff raster format. Finally Geographic Information System (ArcGIS\textregistered \  of ESRI) was used to design the final map (Figure \ref{fig:mass02}). This map (Fig. \ref{fig:mass02}) represents the landslide probability of occurrences in MASS. According to Fig. \ref{fig:mass02} higher probability of landslide occurrence are around 0.8, and are located in the northern slopes of Balsamo Range (western side of MASS), in the eastern slope of San Salvador volcano and Picacho,  along the extension of Mariona Hills  and in the southeastern side, between Ilopango lake and San Jacinto Hill. Areas with lower probability but still important, are located in southern and western side of Nejapa Hill, in north-east of the study area, in northern banks of Ilopango lake, in northeast slope of San Salvador Volcano, in the surrounding areas of San Jacinto Hill and, in general in southern side of MASS.

@@

In the process of estimating the weights of the model, an heuristic technique was used to obtain a better solution after that a local search was used. The difficulty of this approach is due to intensive computation involved in the estimation of a neural network model takes between 4 and 10 hours. For this reason the statistical significance of the input variables could not be assessed by means boostrapping  (a statistical resampling technique). For all of the above parallel computing must be used rather than serial computing to estimate weights of a neural networks models.

The results obtained through geostatistical methodology met the assumptions for kriging method application. Final map of landslide probability of occurrence (see Fig. \ref{fig:mass02} ) generated by kriging method was the result of multidisciplinary work which implied an intense geological analysis (i.e. MARN inventory of 4792 landslides) together with the use of sophisticated statistical techniques. This landslide hazard map is a user-friendly tool to evaluate landslide risk in MASS and can be useful for many purposes such as territorial planning, prevention and landslide mitigation and so on. 
\begin{center}
  \begin{figure}
   \centering
   \includegraphics[scale=0.73]{MASS_mapa_2}
   \caption{\small{Landslide hazard map MASS where the landslide probability of occurrence is shown.}}
   \label{fig:mass02}
  \end{figure}
 \end{center}
High probability of landslide occurrence in Picacho slope implies very high risk due the proximity of urban areas and due the large volume of mobilized materials. Major et al.\cite{major2004} identified in San Salvador volcano lahars as voluminous as 2 million of $m^{3}$ that can affect large zones of urban areas. Southern areas of MASS also implies a high risk because landslides, similar to Las Colinas landslide which involved a total volume of around 200,000 m3 \cite{evans}, can occur in northern slope of Balsamo range and San Jacinto Hill affecting very crowded urban areas.

The present hazard landslide map is consistent with former landslide hazard maps carried out in the region using approaches conceptually different. Fern\'{a}ndez-Lavado et al.\cite{fernan2008} presented a landslide hazard GIS map of MASS generated through bivariate statistical method considering several conditioning factors showing similar hazard areas. The only differences are located mainly in the Nejapa Hill, San Jacinto Hill where Fern\'{a}ndez-Lavado et al. \cite{fernan2008} identified a slightly higher hazard and the area between San Jacinto Hill and Ilopango Lake where the former study presented a hazard slightly lower. As landslides are an erosional phenomenon, the highest probability areas of landslide occurrence coincides with the areas with highest hazard of erosion in the erosion hazard map of MASS proposed by Ch\'{a}vez et al. \cite{chavez2014a}. There are similarities with susceptibility hazard maps at national scale generated through logistic regression method \cite{garcia2008}, through artificial neural network method \cite{garcia2010} and through Mora-Vahrson method \cite{snet2004}. However the high landslide areas are more extended in these national maps than the present study. These differences could be explained by the working scale, our map is more detailed because is at regional scale. Other possible explanation can be related to the data, present work was built using a more extended landslide catalog and more updated data (e.g. seismic hazard evaluation from Benito et al. \cite{beni2012} than former works.

Rock falls were not included in the catalog; these were very common during 2001 earthquakes in El Balsamo Range \cite{jibson}. On the other hand, the MASS seismic microzonation of seismic site effects will improve the estimation of peak ground acceleration. The local geologic conditions are responsible for the modifications (amplification, frequency content and duration) experimented by seismic motions just before reaching the ground surface, additionally, the geometry of the ground surface also may produce amplification (topographic effects) \cite{aki1993}. These effects can be important in soft soils close to incised rivers and also in San Salvador volcano, Balsamo ranges and hills. Crosta et al. \cite{crosta2005} identified an amplification factor of 1.3-1.4 in the Las Colinas failure area (Balsamo Ranges). All these variables can improve the predictive ability of the model and surely the likelihood of landslides.

\section{Acknowledgment}   
The authors are indebted with MARN who provided the data and Geographic Information System software such as ILWIS and ArcGIS\textregistered \  of ESRI.

>>>>

Throughout the year the country experiences a tropical climate with two seasons: a dry season (November to April) and wet season (May - October). The climate of El Salvador is generally warm. In the dry season there is very little rainfall but during rainy seasons heavy showers take place. Periodically El Salvador is affected by tropical storms and occasionally by hurricanes.

@@

\section{Brief State of Art}
\label{sec:brief}
Since the pioneering work shown in \cite{Carrara1983403}, several mathematical and statistical models have been proposed to model landslide susceptibility: deterministic models (\cite{hessd-10-12643-2013},  \cite{doi:10.1080/19475705.2010.498151}, and \cite{Neu2012511}) and probabilistic models (\cite{Bern198839}, \cite{Chung2003451}, and \cite{doi:10.1080/01431160310001618734}).

Popular classification models have been used such as logistic regression (\cite{akgun2012}, \cite{gaskill}, and \cite{garcia2008}), neural networks (\cite{garcia2010}, \cite{Melchiorre2011410},\cite{Zeng2001374}, \cite{Ermini2005327}, and \cite{Yesilnacar2005251}), and support vector machines (\cite{ballabio2012support} and \cite{tien2012landslide}).

According to \cite{van2006landslide}, the magnitude of a possible
slide is difficult to foresee as it depends on the magnitude of the triggering event and the environmental conditions (e.g., height of water table) at the moment of the event. Because of these complex relationships between the dependent variable and causal factors, and since neural networks are particularly useful for detecting complex non-linear relationships in large datasets, we chose this kind of model, despite the disadvantages such as greater computational burden and tendency to overfit.

\section{Study area}
\label{sec:studyarea}
MASS is formed by 14 municipalities among them is located San Salvador, the capital of the country. It has 587 $km^{2}$ and according to the most recent census \cite{minecon} it has 1,566,629 inhabitans with a density of population of 2669 hab/km2. Geographically, MASS is extended over a flat erosional surface (650-760 m.a.s.l.), gently sloping to the east towards Ilopango Caldera also known as Ilopango Lake. MASS is bordered to the south by the Balsamo Range and San Jacinto Hill, to the west by San Salvador volcano and Mariona Hills (refer to Fig. \ref{fig:mass01}). Human settlement is not restricted to the plain and it spreads up to surrounding heights and volcano flanks.

All rocks exposed in MASS have volcanic origin and consists of intercaled primary and reworked deposits from Late Tertiary to Holocene age \cite{schmidt1975}. The younger ones are known as Tierra Blanca pyroclastic ash deposits and forms a relative continuous layer in the majority of MASS with an average thickness of 4 m \cite{schmidt1975}. These are deposits derived mainly from Ilopango caldera eruptions where they are more than 50 m of thick \cite{schmidt1975}. Tierra Blanca soils is highly susceptible to erosion, one of the most common dangerous effects of the erosion in this area are landslides during heavy rainfall and earthquake ground shaking \cite{schmidt1975}. Ch\'{a}vez et al. \cite{chavez2014a} identified high and moderate erosion hazard in the majority of MASS area. The more recent layer Tierra Blanca, known as Tierra Blanca Joven (TBJ) \cite{hernan2004} presents very poor geotechnical properties, especially when there is an increment of soil moisture (i.e. during rainy seasons). This results in high instability susceptibility (\cite{chavez2014b} and \cite{rolo2004}). Several incised rivers and ravines cross MASS area. Due to the large erosion rates of the area, natural slopes of these watercourses are often close to vertical and can reach highs of over 10 m. Taking into account this property of volcanic soils, usually vertical slopes are cut to urban and road construction. As Bommer \& Rodriguez \cite{bommer} identified in Central America, although such slopes may remain stable for years, they may become unstable, abruptly and totally under the action of heavy rainfall or seismic shaking. Hern\'{a}ndez \cite{hernan2004} carried out a detailed description of this process in TBJ, where these instabilities are very common. 
In the study area fault trends are characterized by steep dip angles (65\textdegree - 90\textdegree). The main fault system is characterized by an east-west trend; this trend is responsible for the steep northern slope of the El Balsamo Range.  The secondary systems are with north, northwest and northeast trends offset the main fault system and are more active in the present tectonic setting \cite{schmidt1975}. However, all the mentioned fault families appear active. Thus the several fault systems have developed at different times, but been repeatedly activated \cite{rymer1987}. Finally there are also several ringlike faults that can be related to subsurface collapse due to volcanic tectonic subsidence \cite{schmidt1975}. The Last eruption of San Salvador volcano was in 1917.  The majority of the magma's total volume (97\%) extruded from several vents in the Northwest flank of the volcano, outside of MASS. However a fraction of MASS is located under the proximal volcanic hazard zone according to Sofield \cite{sofield2004}. In the steep slopes  San Salvador volcano, especially of Picacho, the highest peak of the volcano, located in the eastern side presents a high lahar hazard that threats MASS \cite{major2004}. On the other hand one of the most destructive eruptions in the world history was 430 A.D. TBJ eruption produced by the Ilopango Caldera. According to Dull \cite{dull2004}, that eruption produced total devastation in whole MASS area.

\section{Landslide inventory and data sources of input variables}
\label{sec:landsinvet}
The Landslide inventory was developed by Ministry of Environment and Natural Resources (MARN Spanish acronym) and it was conducted in two stages. The first was carried out, based on fieldwork after the 2001 earthquakes in El Salvador. The second stage consisted of a visual inventory of places where landslides were identified through satellite image from IRS-C sensor (Indian) in the same year.

@@

\begin{enumerate}
\item {\bf Geomorphology:} Refers to landforms that result from lithospheric dynamics of geographic area. This variable was obtained in the project of national land use plan in El Salvador. It is unpublished geomorphologic cartography in digital format provided by MARN.

@@

\item {\bf Geology:} Description of the geology of the area according to digital geological map of  El Salvador (scale 1:100,000) \cite{weber}.

@@

\item {\bf Road distance:} Distance in kilometers to the nearest road. Road map was obtained from digital national map created by Centro Nacional de Registros de la Rep\'{u}blica de El Salvador (CNR).

\item {\bf Fault distance:} Distance in kilometers to the nearest fault. This information was obtained from the according digital geological map of El Salvador (scale 1:100,000) \cite{weber}.

@@

\section{Artificial Neural Network model for discrete choice}
Logistic regression represents a neural network with 
a neuron in the hidden layer, whose output is dichotomous. The following adapted form of the Multilayer Perceptron (MLP) may be used for modeling binary classification problems. In the equation (\ref{eq:ann1}) $x_{k,i}$ are the observed values in the $i$th input variables belonging to the $k$th case. The expressions $w_{j,k}$ and $\lambda_{j}$ showed in the equations (\ref{eq:ann1}) and (\ref{eq:ann3}) are the parameters of the model.

@@

In the equation (\ref{eq:ann3}) $p_{i}$ is the the predicting probability for a network with $ k^{*} $ input characteristics and $ j^{*} $ neurons. In the context of the present paper, $p_{i} $ represents the probability of landslide ocurrence.

@@

In the equation (\ref{eq:scaling}) $z$ is the new variable transformed. $ \bar{x} $ and $\sigma_{x}$ are the mean and the standard deviation of the input variable to transform, respectively.

The method used for estimating the parameters of the model was an Hybrid Method \cite{McNelis2005}. Firstly, a heuristic genetic algorithm, using a package developed in the R statistical software \cite{rproject} specifically developed for this purpose, was used to obtain a good estimation of the parameters of the model. The R package can be accessed from the following Web address:

@@

Once a good estimate was obtained, this was used as the initial guess value for the conjugate gradient method implemented in the R statistical software (optim function), to obtain a better estimation of the parameters of the model.

@@

In standard statistical problems, correlation can be estimated from a scatterplot, when several data pairs ${x, y}$ are available. The spatial correlation between two observations of a variable $z(s)$ at locations $s_{1}$ and $s_{2}$ cannot be estimated because only a single pair is available. To estimate spatial correlation
from observational data, it is necessary to make stationarity assumptions
before we can make any progress. One commonly used form of stationarity
is intrinsic stationarity, which assumes that the process that generated the
samples is a random function $Z(s)$ composed of a mean and residual \cite{bivand2008applied}:

@@

\begin{equation}
\min \sigma^{2}_{e} = E\left[Z(s_{0}) - \hat{Z(s_{0})}\right]^{2}
\label{eq:meankrige}
\end{equation}

\section{Results}
The data were randomly divided into three sub-samples: The first was used for the
estimation of the parameters of the neural network model (training set), the second was used to choose the model with more generalization capability (validation set), and the last was
used to evaluate how well the model is able to generalize data not used in the estimation process (test set).

To determine the number of neurons in the hidden layer, the method of trial and error was used, starting with few neurons and increasing progressively the number of neurons in the hidden layer. Table \ref{annresults}
summarizes how well the models fit on the training, validation, and test sets.

@@

The model with 17 neurons in the hidden layer was chosen because it had the best performance in the validation set. After that a logistic regression model was fitted as a basis of comparison. Table \ref{logisticresults} shows logistic regression results. Clearly, the neural network model chosen has better performance than the logistic regression model.

@@

Developed to perform the previous steps, an R script was developed. This script can be downloaded in:

@@

First of all, one thousand fifty georeferenced points with their probability of occurrence calculated using neural network, were selected randomly for the map generation process leaving the others for validation of the kriging model. The Exploratory analysis showed the presence of spatial correlation. This was done making scatter plot of pairs $Z(s_{i})$ and  $Z(s_{j})$, grouped according to their separation distance. Furthermore, the data did not show the presence of anisotropic effect.

Regarding variogram estimation, the spherical, gaussian and exponential models were proposed. Using cross-validation methodology, the best model was the exponential, since it had an $ R^{2} = 0.42 $ against $ R^{2} = 0.40 $ and $ R^{2} = 0.39 $ of the models spherical and gaussian, respectively. Using exponential model, spatial predictions were calculated on a grid of 1500 $\times$ 1500 and exported to Geotiff raster format. Finally, Geographic Information System (ArcGIS\textregistered \  of ESRI) was used to design the final map (Figure \ref{fig:mass02}). This map (Fig. \ref{fig:mass02}) represents the landslide probability of occurrences in MASS. According to Fig. \ref{fig:mass02} higher probability of landslide occurrence are around 0.8. These are located in the northern slopes of Balsamo Range (western side of MASS), in the eastern slope of San Salvador volcano and Picacho,  along the extension of Mariona Hills  and in the southeastern side, between Ilopango lake and San Jacinto Hill. Areas with lower probability, but still important, are located in southern and western side of Nejapa Hill, in north-east of the studied area, in northern banks of Ilopango lake, in northeast slope of San Salvador Volcano, in the surrounding areas of San Jacinto Hill, and in general in southern side of MASS.

@@

In the process of estimating the weights of the model, a heuristic technique was used to obtain a better solution after a local search was used. The difficulty of this approach is due to the intensive computation involved in the estimation of a neural network model, which takes between 4 and 10 hours. For this reason the statistical significance of the input variables could not be assessed by means of bootstrapping (a statistical resampling technique). For all of the above, parallel computing must be used rather than serial computing to estimate weights of the neural networks models.

The results obtained through geostatistical methodology met the assumptions for kriging method application. Final map of landslide probability of occurrence (see Fig. \ref{fig:mass02} ) generated by kriging method was the result of multidisciplinary work which implied an intense geological analysis (i.e. MARN inventory of 4792 landslides), together with the use of sophisticated statistical techniques. This landslide hazard map is a user-friendly tool to evaluate landslide risk in MASS and can be useful for many purposes such as territorial planning, prevention and landslide mitigation and so on. 
\begin{center}
  \begin{figure}
   \centering
   \includegraphics[scale=0.73]{MASS_mapa_2}
   \caption{\small{Landslide hazard map MASS where the landslide probability of occurrence is shown.}}
   \label{fig:mass02}
  \end{figure}
 \end{center}
High probability of landslide occurrence in Picacho slope implies very high risk due to the proximity of urban areas and due to the large volume of mobilized materials. Major et al.\cite{major2004} identified in San Salvador volcano lahars as voluminous as 2 million of $m^{3}$ that can affect large zones of urban areas. Southern areas of MASS also have a high risk because landslides, similar to Las Colinas landslide which involved a total volume of around 200,000 $m^{3}$ \cite{evans}, can occur in northern slope of Balsamo range and San Jacinto Hill, affecting very crowded urban areas.

The present hazard landslide map is consistent with former landslide hazard maps carried out in the region using approaches conceptually different. Fern\'{a}ndez-Lavado et al.\cite{fernan2008} presented a landslide hazard GIS map of MASS generated through bivariate statistical method considering several conditioning factors showing similar hazard areas. The only differences are located mainly in the Nejapa Hill and San Jacinto Hill where Fern\'{a}ndez-Lavado et al. \cite{fernan2008} identified a slightly higher hazard and the area between San Jacinto Hill and Ilopango Lake where the former study presented a hazard slightly lower. Since landslides are an erosional phenomenon, the highest probability areas of landslide occurrence coincide with the areas with highest hazard of erosion in the erosion hazard map of MASS proposed by Ch\'{a}vez et al. \cite{chavez2014a}. There are similarities with susceptibility hazard maps at national scale generated through logistic regression method \cite{garcia2008}, through artificial neural network method \cite{garcia2010}, and through Mora-Vahrson method \cite{snet2004}. However the high landslide areas are more extended in these national maps than the present study. These differences could be explained by the working scale. Our map is more detailed because is at regional scale. Other possible explanation can be related to the data. Present work was built using a more extended landslide catalog and more up-to-date data (e.g. seismic hazard evaluation from Benito et al. \cite{beni2012}) than former works.

Rock falls were not included in the catalog; these were very common during 2001 earthquakes in El Balsamo Range \cite{jibson}. On the other hand, the MASS seismic microzonation of seismic site effects will improve the estimation of peak ground acceleration. The local geologic conditions are responsible for the modifications (amplification, frequency content and duration) experimented by seismic motions just before reaching the ground surface. Additionally, the geometry of the ground surface also may produce amplification (topographic effects) \cite{aki1993}. These effects can be important in soft soils close to incised rivers and also in San Salvador volcano, Balsamo ranges and hills. Crosta et al. \cite{crosta2005} identified an amplification factor of 1.3-1.4 in the Las Colinas failure area (Balsamo Ranges). All these variables can improve the predictive ability of the model and the likelihood of landslides.

\section{Acknowledgment}   
The authors are indebted with MARN for providing the data and Geographic Information System software such as ILWIS and ArcGIS\textregistered \  of ESRI.
